---
title: "ST2195 Coursework Project R Question 4"
author: "Jessica Lawrence Gunawan - 210421794"
date: "2023-01-02"
output: html_document
---

# ST2195 Coursework

## Question 4 : Can you detect cascading failures as delays in one airport create delays in others?

```{r}
# Set working directory

setwd("~/Desktop/ST2195 Coursework Project")
```

```{r}
# Import and load necessary libraries

library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
# Load the datasets

flight2005 <- read.csv("2005.csv", header = TRUE)
flight2006 <- read.csv("2006.csv", header = TRUE)

airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
plane_data <- read.csv("plane-data.csv", header = TRUE)
```

```{r}
# Row binding the flights data in 2005 and 2006

flights <- rbind(flight2005, flight2006)
```

```{r}
# Drop rows with missing values

flights <- flights %>%
  drop_na(DepTime, ArrTime, DepDelay, ArrDelay)
```

```{r}
# Re-check for any missing values

colSums(is.na(flights))
```

```{r}
# Merge between flights and airports

cf <- left_join(flights, airports, by = c("Origin" = "iata"))

cf
```

```{r}
# Querying necessary columns to analyse cascading failures

cf <- cf %>% 
  select(Year, Month, DayofMonth, DayOfWeek, DepTime, CRSDepTime, ArrTime, CRSArrTime, TailNum, ArrDelay, DepDelay, Origin, Dest, LateAircraftDelay, airport)

cf
```

### Explanation

To analyze cascading failures, we will identify whether a delay in departure time of a plane will lead to delay for the next schedule of that plane. There are cascading failures if the plane departs late for its next schedule.

We will take 3 samples to prove whether cascading failures exists or not. Here, we will use 'LateAircraftDelay' variable as it measures the amount of time by which a flight is delayed due to the previous flight's late arrival. This delay occurs because the aircraft used for the current scheduled flight is the same as the one used for the previous flight.

```{r}
# Select rows with LateAircraftDelay between 0 and 120. We only take delays between 0 and 120 minutes as further delays might result in the following flight being canceled. Then, sort the selected rows by LateAircraftDelay in descending order

cf_lateaircraftdelay <- cf %>%
  filter(LateAircraftDelay > 0 & LateAircraftDelay < 120) %>%
  arrange(desc(LateAircraftDelay))

cf_lateaircraftdelay
```

We will choose three samples randomly from above. The chosen three samples are :
- "Year" = 2006, "Month" = 12, "DayofMonth" = 4, "TailNum" = "N664MS"
- "Year" = 2005, "Month" = 2, "DayofMonth" = 25, "TailNum" = "N809MD"
- "Year" = 2005, "Month" = 3, "DayofMonth" = 26, "TailNum" = "N566JB"

```{r}
# First sample

cf_N664MS <- cf %>% 
  filter(Year == 2006, Month == 12, DayofMonth == 4, TailNum == "N664MS") %>% 
  arrange(CRSDepTime)
  
cf_N664MS
```

```{r}
# Plotting a line graph for the first sample

ggplot() +
  geom_line(data = cf_N664MS, aes(y = DepTime, x = airport, color = "DepTime"), group = 1) +
  geom_line(data = cf_N664MS, aes(y = CRSDepTime, x = airport, color = "CRSDepTime"), linetype = "dashed", group = 1) +
  labs(x = "Airport", y = "Departure Time", title = "Delays in One Airport Create Delays in Others") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  scale_linetype_manual("", values = c("CRSDepTime" = "dashed", "DepTime" = "solid")) +
  scale_color_manual(name = "Lines", values = c("CRSDepTime" = "orange", "DepTime" = "blue")) +
  scale_x_discrete(limits = c("Lambert-St Louis International", "Nashville International", "Chicago O'Hare International", "Central Illinois Regional", "Evansville Regional", "Des Moines International"))
```

```{r}
# Second sample

cf_N809MD <- cf %>% 
  filter(Year == 2005, Month == 2, DayofMonth == 25, TailNum == "N809MD") %>% 
  arrange(CRSDepTime)
  
cf_N809MD
```

```{r}
# Plotting a line graph for the second sample

ggplot() +
  geom_line(data = cf_N809MD, aes(y = DepTime, x = airport, color = "DepTime"), group = 1) +
  geom_line(data = cf_N809MD, aes(y = CRSDepTime, x = airport, color = "CRSDepTime"), linetype = "dashed", group = 1) +
  labs(x = "Airport", y = "Departure Time", title = "Delays in One Airport Create Delays in Others") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  scale_linetype_manual("", values = c("CRSDepTime" = "dashed", "DepTime" = "solid")) +
  scale_color_manual(name = "Lines", values = c("CRSDepTime" = "orange", "DepTime" = "blue")) +
  scale_x_discrete(limits = c("Ronald Reagan Washington National", "Detroit Metropolitan-Wayne County", "Philadelphia Intl", "Lambert-St Louis International"))
```

```{r}
# Third sample

cf_N566JB <- cf %>% 
  filter(Year == 2005, Month == 3, DayofMonth == 26, TailNum == "N566JB") %>% 
  arrange(CRSDepTime)
  
cf_N566JB
```

