---
title: "ST2195 Coursework Project R Question 2"
author: "Jessica Lawrence Gunawan - 210421794"
date: "2023-01-02"
output: html_document
---

# ST2195 Coursework

## Question 2 : Do older planes suffer more delays?

```{r}
# Set working directory

setwd("~/Desktop/ST2195 Coursework Project")
```

```{r}
# Import and load necessary libraries

library(dplyr)
library(tidyverse)
library(ggplot2)
```

```{r}
# Load the datasets

flight2005 <- read.csv("2005.csv", header = TRUE)
flight2006 <- read.csv("2006.csv", header = TRUE)

airports <- read.csv("airports.csv", header = TRUE)
carriers <- read.csv("carriers.csv", header = TRUE)
plane_data <- read.csv("plane-data.csv", header = TRUE)
```

```{r}
# Row binding the flights data in 2005 and 2006

flights <- rbind(flight2005, flight2006)
```

```{r}
# Drop rows with missing values

flights <- flights %>%
  drop_na(DepTime, ArrTime, DepDelay, ArrDelay)
```

```{r}
# Re-check for any missing values

colSums(is.na(flights))
```

### Assumptions made :

-   'Year' column in flights dataset refers to be the year that the plane was used to fly by passengers
-   'year' column in plane_data dataset refers to be the year that the plane was manufactured

Therefore, age of plane is gotten by subtracting the value in 'Year' column of flight dataset with the value of 'year' column in plane_data dataset

Based on the assumption made above, there are several cases where the age appears to be a negative value which is does not make sense, and hence we will not include them in the analysis

```{r}
# Look at the summary of plane_data dataset

str(plane_data)
```

```{r}
# Merge between flights and plane_data

plane_delay <- left_join(flights, plane_data, by = c("TailNum" = "tailnum"))

plane_delay
```

```{r}
str(plane_delay)
```

```{r}
# Filtering out any invalid values in 'year' column

plane_delay <- plane_delay %>% 
               filter(year != "0000" & year != "None")
```

```{r}
# Converting the manufacturing years from a string data type to an integer data type

plane_delay <- plane_delay[!is.na(plane_delay$year), ]
plane_delay$year <- as.integer(plane_delay$year)

str(plane_delay)
```

```{r}
# Create a new column that indicates the age of plane

plane_delay$plane_age <- plane_delay$Year - plane_delay$year

plane_delay
```

```{r}
# Group the data by age of plane and count the number of flights

flights_by_age <- plane_delay %>% 
  group_by(plane_age) %>% 
  summarise(`Total Flights` = n()) %>% 
  rename(`Age of Plane` = plane_age)
```

```{r}
# Sort values by Age of Plane in ascending order

flights_by_age <- flights_by_age %>%
  arrange(`Age of Plane`)

flights_by_age
```

```{r}
# Querying necessary data to analyse average delay and total delayed flights based on the age of plane

plane_age_mean <- plane_delay %>%
  filter(between(plane_age, 0, 50), ArrDelay > 0) %>%
  group_by(plane_age) %>%
  summarise("Average Arrival Delay" = mean(ArrDelay, na.rm = TRUE),
            "Total Delayed Flights" = n()) %>%
  rename("Age of Plane" = plane_age)

plane_age_mean
```

```{r}
# Create scatter plot for analyzing average delay based on the age of plane
ggplot(data = plane_age_mean) +
  geom_point(aes(x = `Age of Plane`, y = `Average Arrival Delay`)) + 
  geom_smooth(aes(x = `Age of Plane`, y = `Average Arrival Delay`), method = "lm", se = FALSE, color = "red") +
  labs(x = "Age of Plane", y = "Average Arrival Delay (min)", title = "Average Delay based on Age of Plane") +
  theme_classic() +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```

It can be seen from the data above, that the distribution of the total flights based on the age of plane is not equal. The sample size for older planes are relatively smaller than the younger planes. Therefore, it is better to check futher on the distribution of the total flights based on the age of plane.

```{r}
# Plot a histogram of total delayed flights versus age of plane

ggplot(plane_age_mean, aes(x = `Age of Plane`, weight = `Total Delayed Flights`)) +
  geom_histogram(bins = 50, color = "black", fill = "blue") +
  labs(title = "Histogram of Total Flights versus Age of Plane", x = "Age of Plane", y = "Total Delayed Flights")
```

From the histogram above, we can conclude that the sample size for age below 23 is relatively bigger. Hence, we will plot another scatter plot for age between 0 to 23 i.e. age that has bigger sample size. This is so that we can get more accurate result

```{r}
# Observing average arrival delay based on the age of plane between 0 and 23

plane_age_mean_2 <- plane_delay %>%
  filter(between(plane_age, 0, 23), ArrDelay > 0) %>%
  group_by(plane_age) %>%
  summarise("Average Arrival Delay" = mean(ArrDelay, na.rm = TRUE),
            "Total Delayed Flights" = n()) %>%
  rename("Age of Plane" = plane_age)

plane_age_mean_2
```

```{r}
# Create a scatter plot for analyzing average delay based on the age of plane between 0 and 23

ggplot(data = plane_age_mean_2) +
  geom_point(aes(x = `Age of Plane`, y = `Average Arrival Delay`)) + 
  geom_smooth(aes(x = `Age of Plane`, y = `Average Arrival Delay`), method = "lm", se = FALSE, color = "red") +
  labs(x = "Age of Plane", y = "Average Arrival Delay (min)", title = "Average Delay based on Age of Plane") +
  theme_classic() +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```

```{r}
# Querying necessary data to analyse percentage of delayed flights based on the age of plane

plane_age_percentage <- plane_delay %>% 
  filter(between(plane_age, 0, 50)) %>%
  group_by(plane_age) %>%
  summarise('Total Delayed Flights' = sum(ArrDelay > 0, na.rm = TRUE),
            'Total Flights' = n()) %>% 
  mutate('% of Delayed Flights' = (`Total Delayed Flights` / `Total Flights`) * 100)

plane_age_percentage
```

```{r}
# Create a scatter plot for analyzing % of delayed flights based on the age of plane

ggplot(data = plane_age_percentage) +
  geom_point(aes(x = plane_age, y = `% of Delayed Flights`)) + 
  geom_smooth(aes(x = plane_age, y = `% of Delayed Flights`), method = "lm", se = FALSE, color = "red") +
  labs(x = "Age of Plane", y = "% of Delayed Flights", title = "Percentage of Delayed Flights based on Age of Plane") +
  theme_classic() +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```

```{r}
# Observing percentage of delayed flights based on the age of plane between 0 and 23

plane_age_percentage_2 <- plane_delay %>% 
  filter(between(plane_age, 0, 23)) %>%
  group_by(plane_age) %>%
  summarise('Total Delayed Flights' = sum(ArrDelay > 0, na.rm = TRUE),
            'Total Flights' = n()) %>% 
  mutate('% of Delayed Flights' = (`Total Delayed Flights` / `Total Flights`) * 100)

plane_age_percentage_2
```

```{r}
# Create a scatter plot for analyzing % of delayed flights based on the age of plane between 0 and 23

ggplot(data = plane_age_percentage_2) +
  geom_point(aes(x = plane_age, y = `% of Delayed Flights`)) + 
  geom_smooth(aes(x = plane_age, y = `% of Delayed Flights`), method = "lm", se = FALSE, color = "red") +
  labs(x = "Age of Plane", y = "% of Delayed Flights", title = "Percentage of Delayed Flights based on Age of Plane") +
  theme_classic() +
  theme(axis.text.x = element_text(vjust = 0.5),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 
```
